A mechanization of semi-simplicial and semi-cubical sets in the calculus of inductive constructions

### 2022-08-29 paper/paper.tex
Proposal for beginning of overture.

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  We present a parametricity-based construction of semi-simplicial sets, that naturally generalizes to semi-cubical sets as well, in the calculus of inductive constructions. The accompanying artifact is an axiom-less~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

\section{Overture}
Our contribution is the following. We have modelled the core of semi-simplicial sets in a way that naturally generalizes to semi-cubical sets in type theory, and we call our generalization \emph{$\nu$-sets}. In contrast to previous efforts that use a fibered representation, we use an indexed representation. Our work is, in principle axiom-free, although the accompanying artifact uses functional extensionality, and this requires disappears if $\nu$ is finite.

@@

Cubical type theory~\cite{Bezem14}~\cite{Cohen16}~\cite{Angiuli17} is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. The most significant implementation of cubical type theory in the real world is a branch of Agda, known as \emph{Cubical Agda}~\cite{Vezzosi21}. Cubical type theory is usually modeled as cubical sets, in the set-theoretic setting.

>>>>

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  We present a parametricity-based construction of semi-simplicial sets and semi-cubical sets in indexed form, in the calculus of inductive constructions. The accompanying artifact is an axiom-less~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

\section{Overture}

In the context of typed functional programming, Reynolds'
parametricity~\cite{Reynolds72} interprets types as relations
characterizing the observational equality of programs of this type.
Parametricity can be iterated and it has been noted that iterated
Reynolds' parametricity has a cubical
flavor~\cite{AltenkirchKaposi15,JohannSojakova17,Moulin16,...}. We obtain a
\emph{unary} variant of Reynolds' \emph{binary} parametricity by using
predicates instead of relations, in which case we obtain a form of
(Kreisel) realizability~\cite{Moulin16,BernardyMoulin12}. It has then be
noted that iterated unary parametricity has an augmented simplicial
flavor\footnote{Observation from Hugo Moeneclaey}.

We exploit the connection between iterated unary parametricity and
augmented simplicial sets and between binary parametricity and cubical
sets to present a uniform construction of both semi-simplicial
semi-cubical sets obtained by applying simple rules iterating
parametricity.  In contrast to the presheaf presentation of
semi-simplicial and semi-cubical sets, our uniform construction, which
we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of
having the set of semi-simplices or semi-cubes in dimension $n+1$
\emph{fibered} over the set of semi-simplices or semi-cubes in
dimension $n$, we consider families of semi-simplices and semi-cubes
indexed over their faces.

\iffalse
Our contribution is the following. We have modelled the core of semi-simplicial sets in a way that naturally generalizes to semi-cubical sets in type theory, and we call our generalization \emph{$\nu$-sets}. In contrast to the presheaf presentation of semi-cubical -based where the set of  that use a fibered representation, we use an indexed representation. Our work is, in principle axiom-free, although the accompanying artifact uses functional extensionality, and this requires disappears if $\nu$ is finite.

@@

Cubical type theory~\cite{Bezem14}~\cite{Cohen16}~\cite{Angiuli17} is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. The most significant implementation of cubical type theory in the real world is a branch of Agda, known as \emph{Cubical Agda}~\cite{Vezzosi21}. Cubical type theory is usually modeled as cubical sets, in the set-theoretic setting.
\fi
